The Unicode Cookbook \linebreak for Linguists

### 2018-01-13 book/chapters/orthography_profiles.tex
deals with chapters 2 and 3 -- wip

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
(using NFD normalization), delimited here by space. Unicode tokenization is
required because sequences of code points can differ in their visual and logical
orders. For example, <õ̰> is ambiguous to whether it is the sequence of <o> +
<\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
computationally they are different. Unicode normalization should be applied to
this string to reorder the code points into a canonical order, allowing the data
to be treated canonically equivalently for search and comparison.

In (2), the
Unicode code points have been logically normalized and visually organized into
grapheme clusters, as specified by the Unicode Standard. The combining character
sequence <õ̰> is normalized and visually grouped together. Note that, the
\textsc{modifier letter small h} at \uni{02B0}, is not grouped with any 
other character. This is because it
belongs to Spacing Modifier Letters category in the Unicode Standard. These
characters are underspecified for the direction in which they modify a host
character. For example, can indicate either pre- or post-aspiration (whereas the
nasalization or creaky diacritic is defined in the Unicode Standard to apply to
a specified base character).

@@

Informally, an orthography profile specifies the graphemes (or, in Unicode
parlance, \textsc{tailored grapheme clusters}) that are expected to occur in any
data to be analyzed or checked for consistency. These graphemes are first
identified throughout the whole data (a step which we call
\textsc{tokenization}), and possibly simply returned as such, possibly including
error messages about any parts of the data that are not specified by the
orthography profile. Once the graphemes are identified, they might also be
changed into other graphemes (a step which we call \textsc{transliteration}).
When a grapheme has different possible transliterations, then these differences
should be separated by contextual specification, possibly down to listing
individual exceptional cases.

@@

\subsection*{Implementation}
For the automatic processing of the profiles, the following technical standards
will be expected:

@@

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{should offer user-options} to specify:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} 
		\item \textsc{the name of the column to be used for transliteration} (if any). 
		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally,
        a warning might be given if the chosen string includes characters from
        the orthography itself. 
		\item \textsc{the symbol-string to be inserted for unmatched strings} in the
        tokenized and transliterated output. 
		\item \textsc{the tokenization method}, i.e.~whether the tokenization should
        proceed \textsc{global} or \textsc{linear} (see B6 above). 
		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile
        should use \textsc{NFC} or \textsc{NFD}. 
    \end{enumerate}
	\item \textsc{might offer user-options}:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{5} 
		\item \textsc{to assist in the ordering of the graphemes.} In our experience, it
        makes sense to apply larger graphemes before shorter graphemes, and to
        apply graphemes with context before graphemes without context. Further,
        frequently relevant rules might be applied after rarely relevant rules
        (though frequency is difficult to establish in practice, as it depends
        on the available data). Also, if this all fails to give any decisive
        ordering between rules, it seems useful to offer linguists the option to
        reverse the ordering from any manual specified ordering, because
        linguists tend to write the more general rule first, before turning to
        exceptions or special cases. 
		\item \textsc{to assist in dealing with upper and lower case characters.} It
        seems practical to offer some basic case matching, so characters like
        <a> and <A> are treated equally. This will be useful in many concrete
        cases, although the user should be warned that case matching does not
        function universally in the same way across orthographies. Ideally,
        users should prepare orthography profiles with all lowercase and
        uppercase variants explicitly mentioned, so by default no case matching
        should be performed. 
		\item \textsc{to treat the profile literal}, i.e.~to not interpret regular
        expression metacharacters. Matching graphemes literally often leads to
        strong speed increase, and would allow users to not needing to worry
        about escaping metacharacters. However, in our experience all actually
        interesting use-cases of orthography profiles include some contexts,
        which automatically prevents any literal interpretation.
    \end{enumerate}
	\item \textsc{should return the following information} to the user:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{8} 
		\item \textsc{the original text-strings} to be processed in the specified
        unicode normalization, i.e.~in either NFC or NFD as specified by the
        user. 
		\item \textsc{the tokenized strings}, with additionally any transliterated
        strings, if transliteration is requested. 
		\item \textsc{a survey of all errors encountered}, ideally both in which
        text-strings any errors occurred and which characters in the
        text-strings lead to errors. 
		\item \textsc{a reordered profile}, when any automatic reordering is offered 
	\end{enumerate}
\end{enumerate}

>>>>

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
(using NFD normalization), delimited here by space. Unicode tokenization is
required because sequences of code points can differ in their visual and logical
orders. For example, <õ̰> is ambiguous to whether it is the sequence of <o> +
<\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
computationally they are different (see Sections \ref{pitfall-ipa-homoglyphs} and \ref{pitfall-homoglyphs-in-IPA} ). 
Unicode normalization should be applied to
this string to reorder the code points into a canonical order, allowing the data
to be treated canonically equivalently for search and comparison.

In (2), the
Unicode code points have been logically normalized and visually organized into
grapheme clusters, as specified by the Unicode Standard. The combining character
sequence <õ̰> is normalized and visually grouped together. Note that, the
\textsc{modifier letter small h} at \uni{02B0}, is not grouped with any 
other character. This is because it
belongs to Spacing Modifier Letters category in the Unicode Standard. These
characters are underspecified for the direction in which they modify a host
character. For example, it can indicate either pre- or post-aspiration (whereas the
nasalization or creaky diacritic is defined in the Unicode Standard to apply to
a specified base character).

@@

Informally, an orthography profile specifies the graphemes (or, in Unicode
parlance, \textsc{tailored grapheme clusters}) that are expected to occur in any
data to be analyzed or checked for consistency. These graphemes are first
identified throughout the whole data (a step which we call
\textsc{tokenization}), and simply returned as such, possibly including
error messages about any parts of the data that are not specified by the
orthography profile. Once the graphemes are identified, they might also be
changed into other graphemes (a step which we call \textsc{transliteration}).
When a grapheme has different possible transliterations, then these differences
should be separated by contextual specification, possibly down to listing
individual exceptional cases.

@@

\subsection*{Implementation}
For the automated processing of the profiles, the following technical standards
will be expected:

@@

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{should offer user-options} to specify:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} 
		\item \textsc{the name of the column to be used for transliteration} (if any). 
		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally,
        a warning might be given if the chosen string includes characters from
        the orthography itself. 
		\item \textsc{the symbol-string to be inserted for unmatched strings} in the
        tokenized and transliterated output. 
		\item \textsc{the tokenization method}, i.e.~whether the tokenization should
        proceed \textsc{global} or \textsc{linear} (see B6 above). 
		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile
        should use \textsc{NFC} or \textsc{NFD}. 
    \end{enumerate}
	\item \textsc{might offer user-options}:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{5} 
		\item \textsc{to assist in the ordering of the graphemes.} In our experience, it
        makes sense to apply larger graphemes before shorter graphemes, and to
        apply graphemes with context before graphemes without context. Further,
        frequently relevant rules might be applied after rarely relevant rules
        (though frequency is difficult to establish in practice, as it depends
        on the available data). Also, if this all fails to give any decisive
        ordering between rules, it seems useful to offer linguists the option to
        reverse the ordering from any manual specified ordering, because
        linguists tend to write the more general rule first, before turning to
        exceptions or special cases. 
		\item \textsc{to assist in dealing with upper and lower case characters.} It
        seems practical to offer some basic case matching, so characters like
        <a> and <A> are treated equally. This will be useful in many concrete
        cases, although the user should be warned that case matching does not
        function universally in the same way across orthographies. Ideally,
        users should prepare orthography profiles with all lowercase and
        uppercase variants explicitly mentioned, so by default no case matching
        should be performed. 
		\item \textsc{to treat the profile literally}, i.e.~to not interpret regular
        expression metacharacters. Matching graphemes literally often leads to
        strong speed increase, and would allow users to not needing to worry
        about escaping metacharacters. However, in our experience all actually
        interesting use-cases of orthography profiles include some contexts,
        which automatically prevents any literal interpretation.
    \end{enumerate}
	\item \textsc{should return the following information} to the user:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{8} 
		\item \textsc{the original text-strings} to be processed in the specified
        unicode normalization, i.e.~in either NFC or NFD as specified by the
        user. 
		\item \textsc{the tokenized strings}, with additionally any transliterated
        strings, if transliteration is requested. 
		\item \textsc{a survey of all errors encountered}, ideally both in which
        text-strings any errors occurred and which characters in the
        text-strings lead to errors. 
		\item \textsc{a reordered profile}, when any automatic reordering is offered 
	\end{enumerate}
\end{enumerate}
